---
title: "Seasonal population distribution"
format: html
editor: visual
---

## Proportion of population by season

The goal is to use eBird status and trend results to calculate the proportion of a species population that is covered by a given monitoring program.

-   seasonal population distribution calculated using

    -   eBird relative abundance maps by season

    -   Boreal Avian Modeling estimates of breeding season population for some Northern species that may be poorly covered by eBird.

-   using a 1-degree lat x long grid for now, but worth re-visiting.

-   method assumes a relatively stationary population during a survey-season (breeding season or non-breeding season, but not applicable to migration monitoring yet)

-   method also assumes that the current distribution/abundance map encompasses the species distribution in 1966.

    -   Note: If the species' range has shifted, areas with monitoring data 50-years ago but no modeled abundance today will be difficult to deal with...

-   starting with BBS survey information, but anticipating a potential application to Christmas Bird Count, eBird trends, and other surveys.

-   Will require a clear taxonomic alignment across various surveys

## downloading eBird seasonal abundance

```{r}
library(ebirdst)
library(tidyverse)
library(bbsBayes2)
library(terra)
library(sf)

```

##Example extraction for Connecticut Warbler

Download species data and load seasonal mean abundance

```{r}

species <- "Connecticut Warbler"

path <- ebirdst_download(species)
abd_seasonal_percent <- load_raster(path, product = "percent-population", 
                                 period = "seasonal", #metric = "mean", 
                                 resolution = "hr") #3km high resolution
abd_breeding_percent <- abd_seasonal_percent$breeding

```


Calculate the mean relative abundance within each stratum. First load the degree-block stratification map, and adjust the crs to match the eBird map.

```{r}

latlong_orig <- load_map("latlong") 

latlong <- load_map("latlong")%>% #loading the 1-degree cell strata
  st_transform(st_crs(abd_breeding_percent)) 

percent_strata <- extract(abd_breeding_percent,
                          latlong,
                          fun = sum,
                          na.rm = TRUE) %>% 
  mutate(strata_name = latlong$strata_name)

map_percent_strata <- latlong_orig %>% 
  inner_join(.,percent_strata,
             by = "strata_name") %>% 
  filter(!is.na(breeding),
         breeding > 0)

map <- ggplot()+
  geom_sf(data = map_percent_strata,
          aes(fill = breeding))

map

```

